Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2022-09-26 paper/paper.tex
Finish paper

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistics with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
\end{abstract}

@@

\listoftodos{}

@@

In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures have been verified in Isabelle/HOL such as a procedure for presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices speed by going through the simplifier or soundness by trusting generated code.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness because the certificates are replayed through Isabelle's inference kernel.

The focus of this paper is another ubiquitous structure in mathematics, namely sets.
In particular, we consider a ground fragment of set theory which \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} call multi-level syllogistics with singleton (\MLSS{}).
The fragment includes the usual set operations of union, intersection, difference, membership, equality and, in addition, it allows the construction of singleton sets.
Due to the omnipresence of sets, we think that automation for this fragment will benefit the users of Isabelle, especially when integrated into the simplifier like the aforementioned order solver.
We also expect that the generation of certificates will be straightforward since \MLSS{} admits a tableau calculus.

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is key for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

All in all, the abstract specification of the decision procedure provides a stepping stone towards a refinement to an executable specification from which code can be generated.

\subsection{Related Work}
AFP entry~\cite{hybrid_logic_afp} and article~\cite{hybrid_logic} and workshop paper in Isabelle 2022 workshop.
A Verified Proof of Craig Interpolation for Basic Modal Logic via Tableaux in Lean.
A Tableau for Ecumenical Propositional Logic.
Automatization of the semantic tableau method.
Formalisation de la logique temporelle dans le coq.
Procedure was later generalised to cartesian product etc.
Procedure was also generate to add monotone functions.
\todo[inline]{Look at all articles of From}

First time MLSS~\cite{mlss_first}.
MLSS is in NP~\cite{mlss_np}.
First MLSS Tableau~\cite{mlss_first_tableau}.
Beckert~\cite{tableau_quantifier_free}, Diploma~\cite{tableau_quantifier_free_diploma}.
Tableau-based MLSS with functions and quantification~\cite{mlss_functions_quantification}.
MLSS with monotone functions~\cite{mlss_monotone_functions}, reduction to MLSS.
Another generalisation with cartesion product and some map constructs~\cite{mlss_cartesian_map}.

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.

We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the space of total functions from \lstinline!'a! to \lstinline!'b!.

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set is determined with \lstinline!card! and \lstinline!f ` A! denotes the image of \lstinline!A! under \lstinline!f!.

Analogously to sets, we use \lstinline!'a list! to describe the type of lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.

We remark that \textbf{iff} is equivalent to \lstinline!$=$! on type of booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
The other important constant of the meta-logic, apart from \lstinline!$\equiv$!, is the meta-implication \lstinline!$\Longrightarrow$!.
A chain of implications
\begin{lstlisting}
  A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C
\end{lstlisting}
can be denoted more concisely by
\begin{lstlisting}
  $\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C.
\end{lstlisting}

\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \MLSS{}, we have the type of set terms which is the disjoint union of the empty set and variables as well as the operations union, intersection, set difference, and the singleton set represented by the constructor \lstinline!Single!.
We keep the type of variables abstract by making it a parameter of the set term data type.
The only restriction on the type of variables is that it needs to be infinite.
Isabelle's data type package automatically defines a function that gives us the set of variables in a set term, which we name \lstinline!vars!.
In what follows, we will overload the function \lstinline!vars! to also work on set atoms, formulae, and branches.
\begin{lstlisting}
datatype (vars: 'a) pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\unionS$ 'a pset_term
| 'a pset_term $\interS$ 'a pset_term
| 'a pset_term $\diffS$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype (vars: 'a) pset_atom =
  'a pset_term $\inS$ 'a pset_term
| 'a pset_term $\eqS$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\subseteqS$! and enumerate finite sets: \lstinline!s $\subseteqS$ t! is equivalent to \lstinline!s $\unionS$ t $\eqS$ t! and a finite set of elements \lstinline!{t$_\text{1}$,$\ldots$,t$_\text{k}$}! can be expressed by \lstinline!Single t$_1$ $\unionS$ $\ldots$ $\unionS$ Single t$_k$!.

@@

Similarly to the function \lstinline!vars!, the automatically defined function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! retrieves all set atoms in a formula.
We combine these functions to extract all the variables occuring in a set formula.
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

In addition to the variable that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
Like \lstinline!vars!, we overload the constant \lstinline!subterms!.
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}

Finally, we consider the subformulae of a formula and define a function that computes them.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}
The \textit{literals} are a type of subformulae of particular importance so we define a predicate for them.
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
\subsection{Semantics}
Formulae in \MLSS{} contain arbitrarily nested set terms.
We account for this nesting by basing the semantics on the von Neumann hierarchy $\mathcal{V}$ of sets that is defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded meaning that there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitiation is provided by Lawrence in an AFP entry~\cite{zfc_in_hol_afp}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
Two articles~\cite{zfc_in_hol1,zfc_in_hol2} have been written that provide further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class.
      This class only constitutes a set if the class is small enough.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a class is small.
  \item The function \lstinline!vset :: 'a set $\Rightarrow$ V! turns a class into a set of type \lstinline!V! if it is small. 
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$ so it is denoted by \lstinline!0 :: V!.
\end{itemize}

@@

For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
The implementation of $\models$ coincides with the interpretation function of \citeauthor{lqe}~\cite{lqe}, so omit it for brevity.
As usual, a formula \lstinline!$\phi$! is called \textit{satisfiable} if there exists a model \lstinline!$M$! with \lstinline!$M$ $\models$ $\phi$!.
Otherwise, \lstinline!$\phi$! is called \textit{unsatisfiable}.

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau. 
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}
We now lift the functions that extract the variables respectively subterms of formulae to branches.
\begin{lstlisting}
definition subterms ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms b $\equiv$ $\bigcup$(subterms ` set b)

definition vars :: 'a branch $\Rightarrow$ 'a set 
  vars b $\equiv$ $\bigcup$(vars ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}

@@

    \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! &&
    \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows all linear expansion rules.
Note that, in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occuring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\begin{lstlisting}
  [&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]
\end{lstlisting}
is saturated --- neither linear nor branching rules apply --- and open but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

@@

Challenges in obtaining an executable specification: membership cycle test unnecessary, fast branch membership test, proof reconstruction.
Additional constructs such as finiteness and pairs.
Optimisations for the algorithm.

>>>>

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
\end{abstract}

@@

%%%%%%%%%\listoftodos{}

@@

In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures have been verified in Isabelle/HOL such as a procedure for Presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices speed by going through the simplifier or soundness by trusting generated code.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness as the certificates are replayed through Isabelle's inference kernel.

The focus of this paper is another ubiquitous structure in mathematics, namely sets.
In particular, we consider a ground fragment of set theory which \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} call multi-level syllogistic with singleton (\MLSS{}).
The fragment includes the usual set operations of union, intersection, difference, membership, equality and, in addition, it allows the construction of singleton sets.
Due to the omnipresence of sets, we think that automation for this fragment will benefit the users of Isabelle, especially when integrated into the simplifier like the aforementioned order solver.
We also expect that the generation of certificates will be straightforward since \MLSS{} admits a tableau calculus.

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

\subsection{Related Work}
Since the literature on decidable fragments of set theory is vast, we only focus on \MLSS{} here.
The fragment was first shown to be decidable by \citeauthor{mlss_first}~\cite{mlss_first}.
Subsequent work~\cite{mlss_np} found the decision problem to be \textbf{NP}-complete.
To obtain a practical decision procedure, \citeauthor{mlss_first_tableau}~\cite{mlss_first_tableau} proposed a tableau calculus, which was later improved by \citeauthor{tableau_quantifier_free}~\cite{tableau_quantifier_free}.
Both of these procedures construct a model during execution that is used to guide the proof search.
\citeauthor{tableau_quantifier_free} also cover an extension of the calculus with uninterpreted functions, which was revisited by \citeauthor{mlss_quantification}~\cite{mlss_quantification} who avoided the construction of a model during execution.
In this paper, we consider a version of the latter procedure due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} that is specialised to \MLSS{} and where the branching rules of the calculus are set up in a way to guarantee mutual exclusivity of the branches.
Later extensions of the calculus added certain interpreted functions such as monotone functions~\cite{mlss_monotone_functions} and the inverse of a function~\cite{mlss_cartesian_map}.
The latter extension notably includes the Cartesian product.
Those extensions, though, did not improve upon the tableau calculus for \MLSS{}.

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
% Formalisation de la logique temporelle dans le coq.

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
We establish notation and in particular some essential data types together with their primitive operations that are specific to Isabelle/HOL.

We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the space of total functions from type \lstinline!'a! to type \lstinline!'b!.

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set \lstinline!A! is denoted by \lstinline!card A! and the image of \lstinline!A! under \lstinline!f! by \lstinline!f ` A!.

We use \lstinline!'a list! to describe the type of lists, which are constructed using the empty list \lstinline![]! constructor or the infix cons constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.

We remark that $\longleftrightarrow$ is equivalent to \lstinline!$=$! on the type of Booleans \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
Meta-implication is denoted by \lstinline!$\Longrightarrow$! and a chain of implications
\begin{lstlisting}
  A$_\text{1}$ $\Longrightarrow$ $\cdots$ $\Longrightarrow$ A$_\text{k}$ $\Longrightarrow$ C
\end{lstlisting}
can be abbreviated by 
\begin{lstlisting}
  $\llbracket$ A$_\text{1}$;$\,\ldots\,$;A$_\text{k}$ $\rrbracket$ $\Longrightarrow$ C.
\end{lstlisting}

\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \MLSS{}, we have the type of set terms which is the disjoint union of the empty set and variables as well as the operations union, intersection, set difference, and the singleton set represented by the constructor \lstinline!Single!.
We keep the type of variables abstract by making it a parameter of the set term data type.
The only restriction on the type of variables is that it needs to be infinite.
Isabelle/HOL's data type package automatically defines a function that gives us the set of variables in a set term, which we name \lstinline!vars!.
In what follows, we will overload the function \lstinline!vars! to also work on set atoms, formulae, and branches.
\begin{lstlisting}
datatype (vars: 'a) pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\unionS$ 'a pset_term
| 'a pset_term $\interS$ 'a pset_term
| 'a pset_term $\diffS$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype (vars: 'a) pset_atom =
  'a pset_term $\inS$ 'a pset_term
| 'a pset_term $\eqS$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\subseteqS$! and enumerate finite sets: \lstinline!s $\subseteqS$ t! is equivalent to \lstinline!s $\unionS$ t $\eqS$ t! and a finite set of elements \lstinline!{t$_\text{1}$,$\ldots$,t$_\text{k}$}! can be expressed by \lstinline!Single t$_1$ $\unionS$ $\ldots$ $\unionS$ Single t$_k$!.

@@

Like the function \lstinline!vars!, the automatically defined function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! retrieves all set atoms in a formula.
We combine these functions to extract all the variables occurring in a set formula.
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms!.
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of its arguments.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Finally, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}

Lastly, we consider the subformulae of a formula and define a function that computes them.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}
\textit{Literals} are a type of subformulae of particular importance, so we define a predicate for them.
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
\subsection{Semantics}
We base the semantics of \MLSS{} on the von Neumann hierarchy $\mathcal{V}$ of sets that is defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded, i.e.\ there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitation is provided by Lawrence in an entry~\cite{zfc_in_hol_afp} of the \textit{Archive of Formal Proofs}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
Two articles~\cite{zfc_in_hol1,zfc_in_hol2} have been written that provide further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class (relative to \lstinline!V!).
      This class only constitutes a set (relative to \lstinline!V!) if the class is small.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a class is small.
  \item The function \lstinline!vset :: 'a set $\Rightarrow$ V! turns a class into a set of type \lstinline!V! if it is small. 
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: V!.
\end{itemize}

@@

We write \lstinline!M $\models$ $\phi$! for the judgement that the formula $\phi$ holds under the valuation function \lstinline!M!.
The implementation of $\models$ coincides with the interpretation function of \citeauthor{lqe}~\cite{lqe}.
As usual, a formula \lstinline!$\phi$! is called \textit{satisfiable} if there exists a model \lstinline!$M$! with \lstinline!$M$ $\models$ $\phi$!.
Otherwise, \lstinline!$\phi$! is called \textit{unsatisfiable}.

\section{A Tableau Calculus for MLSS}
We formalise the tableau calculus for \MLSS{} as described by \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}.
Inspired by the formalisation of a tableau calculus for hybrid logic by \citeauthor{hybrid_logic_afp}~\cite{hybrid_logic_afp}, we simply use a list to represent a branch of the tableau tree.
Note that formulae are added to the front of the list during branch expansion, so \lstinline!last b! for a branch \lstinline!b! is always the formula that we are trying to disprove with the tableau. 
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}
We now lift the functions that extract the variables respectively subterms of formulae to branches.
\begin{lstlisting}
definition vars :: 'a branch $\Rightarrow$ 'a set 
  vars b $\equiv$ $\bigcup$(vars ` set b)

definition subterms ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms b $\equiv$ $\bigcup$(subterms ` set b)
\end{lstlisting}
In the standard tableau calculus for propositional logic as \citeauthor{tableau}~\cite{tableau} describes it, a branch is called \textit{closed} if it contains both the negation of a formula and the formula itself;
conversely, it is called \textit{open} if it is not closed.
For \MLSS{}, we extend the notion of closedness with three additional rules; the first two are straightforward while the last one states that a branch is closed when the branch contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}

@@

    \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! &&
    \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of \lstinline!l!, i.e.\ the set terms that occur directly under one of the atom constructors \lstinline!$\inS$! or \lstinline!$\eqS$!;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

@@

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rule dealing with double negation.
This rule is required for completeness, though, considering that the branch
\begin{lstlisting}
  [&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]
\end{lstlisting}
is saturated --- neither linear nor branching rules apply --- and open, but there clearly is no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

@@

%Challenges in obtaining an executable specification: membership cycle test unnecessary, fast branch membership test, proof reconstruction.
%Additional constructs such as finiteness and pairs.
%Optimisations for the algorithm.
